---
title: "A Replication of Karlan and List (2007)"
author: "Aashvi"
date: 04/23/2025
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
---



## Introduction 

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).


This project seeks to replicate their results.


## Data

### Description

```{r}
# Load necessary packages
library(haven)    # to read .dta
library(dplyr)    # for data manipulation
library(ggplot2)  # for plots
library(broom)    # for tidy regression output
library(knitr)    # for tables

# Read the data
df <- read_dta("karlan_list_2007.dta")

# Show the first few rows
head(df)
```


:::: {.callout-note collapse="true"}
### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |

::::


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.

```{r}
# T-test: compare means of months since last donation
t_test_mrm2 <- t.test(mrm2 ~ treatment, data = df)
t_test_mrm2
```

The average number of months since last donation is not statistically significantly different between treatment and control groups (p-value > 0.05). This suggests randomization worked well and groups were balanced — supporting the integrity of the experiment.

```{r}
# Linear regression: mrm2 on treatment
lm_mrm2 <- lm(mrm2 ~ treatment, data = df)
summary(lm_mrm2)
```

As expected, the regression confirms the t-test: no significant effect of treatment on months since last donation. Coefficient is small and p-value is large. Again, this supports random assignment.


## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

```{r}
library(ggplot2)

# Calculate proportion of donations by group
df %>%
group_by(treatment) %>%
summarise(gave_rate = mean(gave)) %>%
ggplot(aes(x = factor(treatment, labels = c("Control", "Treatment")),
            y = gave_rate,
            fill = factor(treatment))) +
geom_col() +
labs(title = "Proportion of People Who Donated",
    x = "Group",
    y = "Donation Rate") +
theme_minimal() +
theme(legend.position = "none")
```

```{r}
# T-test for donation outcome
t.test(gave ~ treatment, data = df)
```

```{r}
# Linear regression for same effect
lm_gave <- lm(gave ~ treatment, data = df)
summary(lm_gave)
```

The bar plot and t-test show that the treatment group had a higher donation rate than the control group. The difference is statistically significant (p-value < 0.05), meaning the matching offer increased people's likelihood to donate. This supports the paper’s claim that matching gifts raise response rates.

```{r}
# Probit regression
probit_model <- glm(gave ~ treatment, data = df, family = binomial(link = "probit"))
summary(probit_model)
```

The probit regression confirms the earlier findings: being assigned to the treatment increases the probability of donating. This aligns with Table 3, Column 1 in the original paper.


### Differences between Match Rates

Next, I assess the effectiveness of different sizes of matched donations on the response rate.

```{r}
# Subset to treatment group only
treat_df <- df %>% filter(treatment == 1)

# Create dummy variable for 1:1 match (if neither 2:1 nor 3:1)
treat_df <- treat_df %>%
mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0))
```

```{r}
# 1:1 vs 2:1
t_1v2 <- t.test(gave ~ ratio2, data = treat_df %>% filter(ratio1 == 1 | ratio2 == 1))
t_1v2

# 2:1 vs 3:1
t_2v3 <- t.test(gave ~ ratio3, data = treat_df %>% filter(ratio2 == 1 | ratio3 == 1))
t_2v3
```

The t-tests reveal no statistically significant differences in response rates between 1:1 and 2:1 match groups or 2:1 and 3:1 match groups. This supports the authors’ statement that larger match ratios do not lead to higher likelihood of giving, once any match is offered.

```{r}
# Regress giving on ratio dummies
lm_ratios <- lm(gave ~ ratio1 + ratio2 + ratio3, data = treat_df)
summary(lm_ratios)
```

In the regression, all match ratio dummy variables are statistically insignificant. This confirms that the generosity of the match ratio does not meaningfully affect donation rates, and the mere presence of a match drives the behavioral change.

```{r}
# Raw donation rates by ratio group
treat_df %>%
group_by(ratio) %>%
summarise(response_rate = mean(gave))
```

The raw response rates are also similar across 1:1, 2:1, and 3:1 groups — further reinforcing the conclusion that larger matches don't help more than a basic 1:1 match.


### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

```{r}
# T-test: effect of treatment on donation amount (unconditional)
t.test(amount ~ treatment, data = df)
```

```{r}
# Linear regression: amount ~ treatment
lm_amt_all <- lm(amount ~ treatment, data = df)
summary(lm_amt_all)
```

The average donation amount is slightly higher in the treatment group, but the difference is not statistically significant. This suggests that while more people gave, the average amount given (including non-donors) was not substantially larger.

```{r}
# Filter to only people who donated (amount > 0)
df_donors <- df %>% filter(amount > 0)

# T-test: among donors only
t.test(amount ~ treatment, data = df_donors)
```

```{r}
# Linear regression: amount ~ treatment (among donors)
lm_amt_donors <- lm(amount ~ treatment, data = df_donors)
summary(lm_amt_donors)
```

Among those who donated, the treatment group does not give significantly more than the control group. The treatment coefficient does not have a strong causal interpretation here, as it could reflect selection (different kinds of people chose to give in treatment vs. control).

```{r}
library(ggplot2)

# Add group labels
df_donors <- df_donors %>%
mutate(group = ifelse(treatment == 1, "Treatment", "Control"))

# Plot for Control group
df_donors %>% 
filter(group == "Control") %>%
ggplot(aes(x = amount)) +
geom_histogram(binwidth = 5, fill = "steelblue", alpha = 0.7) +
geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
labs(title = "Control Group: Donation Amounts",
    x = "Amount Donated", y = "Count") +
theme_minimal()
```

```{r}
# Plot for Treatment group
df_donors %>% 
filter(group == "Treatment") %>%
ggplot(aes(x = amount)) +
geom_histogram(binwidth = 5, fill = "seagreen", alpha = 0.7) +
geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
labs(title = "Treatment Group: Donation Amounts",
    x = "Amount Donated", y = "Count") +
theme_minimal()
```

Both groups show right-skewed distributions of donation amounts. The dashed red lines represent the average amount given. There is no dramatic difference in the average donation size between treatment and control.


## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.

### Law of Large Numbers

```{r}
# Set seed for reproducibility
set.seed(123)

# Simulate 10,000 draws from Bernoulli p=0.022 (treatment)
treatment_sim <- rbinom(10000, 1, 0.022)

# Simulate 10,000 draws from Bernoulli p=0.018 (control)
control_sim <- rbinom(10000, 1, 0.018)

# Compute difference at each point
diffs <- treatment_sim - control_sim

# Cumulative average of the differences
cum_avg <- cumsum(diffs) / seq_along(diffs)

# Plot
ggplot(data.frame(index = 1:10000, cum_avg = cum_avg), aes(x = index, y = cum_avg)) +
geom_line(color = "steelblue") +
geom_hline(yintercept = 0.004, linetype = "dashed", color = "red") +
labs(title = "Law of Large Numbers Simulation",
    x = "Number of Observations",
    y = "Cumulative Average of Differences") +
theme_minimal()
```

The plot shows the cumulative average of differences in donation probabilities converging toward 0.004 — the true difference (2.2% - 1.8%). This illustrates the **Law of Large Numbers**: as sample size grows, sample statistics converge to their true values.


### Central Limit Theorem

_to do: Make 4 histograms like those on slide 44 from our first class at sample sizes 50, 200, 500, and 1000 and explain these plots to the reader. To do this for a sample size of e.g. 50, take 50 draws from each of the control and treatment distributions, and calculate the average difference between those draws. Then repeat that process 999 more times so that you have 1000 averages. Plot the histogram of those averages. Comment on whether zero is in the "middle" of the distribution or whether it's in the "tail."_

```{r}
simulate_diffs <- function(n, reps = 1000, p_control = 0.018, p_treatment = 0.022) {
replicate(reps, {
    control <- rbinom(n, 1, p_control)
    treatment <- rbinom(n, 1, p_treatment)
    mean(treatment) - mean(control)
})
}

sizes <- c(50, 200, 500, 1000)
diffs_by_size <- lapply(sizes, simulate_diffs)

# Plot histograms
par(mfrow = c(2, 2))  # 2 rows x 2 columns layout

for (i in 1:4) {
hist(diffs_by_size[[i]], breaks = 30, main = paste("Sample size =", sizes[i]),
    xlab = "Mean difference (treatment - control)", col = "lightblue", border = "white")
abline(v = 0, col = "red", lwd = 2, lty = 2)
}
```

As sample size increases, the distribution of average differences becomes more concentrated around the true mean and more symmetric. With larger samples (like 1000), the distribution approximates a normal curve — demonstrating the **Central Limit Theorem**.




