---
title: "A Replication of Karlan and List (2007)"
author: "Aashvi"
date: today
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
---



## Introduction

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).

_to do: expand on the description of the experiment._

This project seeks to replicate their results.


## Data

### Description
_todo: Read the data into R/Python and describe the data_
```{r}
# Load necessary packages
library(haven)    # to read .dta
library(dplyr)    # for data manipulation
library(ggplot2)  # for plots
library(broom)    # for tidy regression output
library(knitr)    # for tables

# Read the data
df <- read_dta("/home/jovyan/Downloads/SPRING/Marketing Analytics/HW1 MA/MA_demo/mysite/HW/karlan_list_2007.dta")

# Show the first few rows
head(df)
```


:::: {.callout-note collapse="true"}
### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |

::::


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.

_todo: test a few variables other than the key outcome variables (for example, test months since last donation) to see if the treatment and control groups are statistically significantly different at the 95% confidence level. Do each as a t-test and separately as a linear regression, and confirm you get the exact same results from both methods. When doing a t-test, use the formula in the class slides. When doing the linear regression, regress for example mrm2 on treatment and look at the estimated coefficient on the treatment variable. It might be helpful to compare parts of your analysis to Table 1 in the paper. Be sure to comment on your results (hint: why is Table 1 included in the paper)._

```{r}
# T-test: compare means of months since last donation
t_test_mrm2 <- t.test(mrm2 ~ treatment, data = df)
t_test_mrm2
```

The average number of months since last donation is not statistically significantly different between treatment and control groups (p-value > 0.05). This suggests randomization worked well and groups were balanced — supporting the integrity of the experiment.

```{r}
# Linear regression: mrm2 on treatment
lm_mrm2 <- lm(mrm2 ~ treatment, data = df)
summary(lm_mrm2)
```

As expected, the regression confirms the t-test: no significant effect of treatment on months since last donation. Coefficient is small and p-value is large. Again, this supports random assignment.

## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

_todo: make a barplot with two bars. Each bar is the proportion of people who donated. One bar for treatment and one bar for control._

```{r}
library(ggplot2)

# Calculate proportion of donations by group
df %>%
group_by(treatment) %>%
summarise(gave_rate = mean(gave)) %>%
ggplot(aes(x = factor(treatment, labels = c("Control", "Treatment")),
            y = gave_rate,
            fill = factor(treatment))) +
geom_col() +
labs(title = "Proportion of People Who Donated",
    x = "Group",
    y = "Donation Rate") +
theme_minimal() +
theme(legend.position = "none")
```

_todo: run a t-test between the treatment and control groups on the binary outcome of whether any charitable donation was made. Also run a bivariate linear regression that demonstrates the same finding. (It may help to confirm your calculations match Table 2a Panel A.) Report your statistical results and interpret them in the context of the experiment (e.g., if you found a difference with a small p-value or that was statistically significant at some threshold, what have you learned about human behavior? Use mostly English words, not numbers or stats, to explain your finding.)_

```{r}
# T-test for donation outcome
t.test(gave ~ treatment, data = df)
```

```{r}
# Linear regression for same effect
lm_gave <- lm(gave ~ treatment, data = df)
summary(lm_gave)
```

The bar plot and t-test show that the treatment group had a higher donation rate than the control group. The difference is statistically significant (p-value < 0.05), meaning the matching offer increased people's likelihood to donate. This supports the paper’s claim that matching gifts raise response rates.


_todo: run a probit regression where the outcome variable is whether any charitable donation was made and the explanatory variable is assignment to treatment or control. Confirm that your results replicate Table 3 column 1 in the paper._

```{r}
# Probit regression
probit_model <- glm(gave ~ treatment, data = df, family = binomial(link = "probit"))
summary(probit_model)
```

The probit regression confirms the earlier findings: being assigned to the treatment increases the probability of donating. This aligns with Table 3, Column 1 in the original paper.


### Differences between Match Rates

Next, I assess the effectiveness of different sizes of matched donations on the response rate.

```{r}
# Subset to treatment group only
treat_df <- df %>% filter(treatment == 1)

# Create dummy variable for 1:1 match (if neither 2:1 nor 3:1)
treat_df <- treat_df %>%
mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0))
```


_todo: Use a series of t-tests to test whether the size of the match ratio has an effect on whether people donate or not. For example, does the 2:1 match rate lead increase the likelihood that someone donates as compared to the 1:1 match rate? Do your results support the "figures suggest" comment the authors make on page 8?_


```{r}
# 1:1 vs 2:1
t_1v2 <- t.test(gave ~ ratio2, data = treat_df %>% filter(ratio1 == 1 | ratio2 == 1))
t_1v2

# 2:1 vs 3:1
t_2v3 <- t.test(gave ~ ratio3, data = treat_df %>% filter(ratio2 == 1 | ratio3 == 1))
t_2v3
```

The t-tests reveal no statistically significant differences in response rates between 1:1 and 2:1 match groups or 2:1 and 3:1 match groups. This supports the authors’ statement that larger match ratios do not lead to higher likelihood of giving, once any match is offered.


_todo: Assess the same issue using a regression. Specifically, create the variable `ratio1` then regress `gave` on `ratio1`, `ratio2`, and `ratio3` (or alternatively, regress `gave` on the categorical variable `ratio`). Interpret the coefficients and their statistical precision._

```{r}
# Regress giving on ratio dummies
lm_ratios <- lm(gave ~ ratio1 + ratio2 + ratio3, data = treat_df)
summary(lm_ratios)
```

In the regression, all match ratio dummy variables are statistically insignificant. This confirms that the generosity of the match ratio does not meaningfully affect donation rates, and the mere presence of a match drives the behavioral change.


_todo: Calculate the response rate difference between the 1:1 and 2:1 match ratios and the 2:1 and 3:1 ratios.  Do this directly from the data, and do it by computing the differences in the fitted coefficients of the previous regression. what do you conclude regarding the effectiveness of different sizes of matched donations?_

```{r}
# Raw donation rates by ratio group
treat_df %>%
group_by(ratio) %>%
summarise(response_rate = mean(gave))
```

The raw response rates are also similar across 1:1, 2:1, and 3:1 groups — further reinforcing the conclusion that larger matches don't help more than a basic 1:1 match.


### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

_todo: Calculate a t-test or run a bivariate linear regression of the donation amount on the treatment status. What do we learn from doing this analysis?_

```{r}
# T-test: effect of treatment on donation amount (unconditional)
t.test(amount ~ treatment, data = df)
```

```{r}
# Linear regression: amount ~ treatment
lm_amt_all <- lm(amount ~ treatment, data = df)
summary(lm_amt_all)
```

The average donation amount is slightly higher in the treatment group, but the difference is not statistically significant. This suggests that while more people gave, the average amount given (including non-donors) was not substantially larger.


_todo: now limit the data to just people who made a donation and repeat the previous analysis. This regression allows you to analyze how much respondents donate conditional on donating some positive amount. Interpret the regression coefficients -- what did we learn? Does the treatment coefficient have a causal interpretation?_ 

```{r}
# Filter to only people who donated (amount > 0)
df_donors <- df %>% filter(amount > 0)

# T-test: among donors only
t.test(amount ~ treatment, data = df_donors)
```

```{r}
# Linear regression: amount ~ treatment (among donors)
lm_amt_donors <- lm(amount ~ treatment, data = df_donors)
summary(lm_amt_donors)
```

Among those who donated, the treatment group does not give significantly more than the control group. The treatment coefficient does not have a strong causal interpretation here, as it could reflect selection (different kinds of people chose to give in treatment vs. control).


_todo: Make two plot: one for the treatment group and one for the control. Each plot should be a histogram of the donation amounts only among people who donated. Add a red vertical bar or some other annotation to indicate the sample average for each plot._

```{r}
library(ggplot2)

# Add group labels
df_donors <- df_donors %>%
mutate(group = ifelse(treatment == 1, "Treatment", "Control"))

# Plot for Control group
df_donors %>% 
filter(group == "Control") %>%
ggplot(aes(x = amount)) +
geom_histogram(binwidth = 5, fill = "steelblue", alpha = 0.7) +
geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
labs(title = "Control Group: Donation Amounts",
    x = "Amount Donated", y = "Count") +
theme_minimal()
```

```{r}
# Plot for Treatment group
df_donors %>% 
filter(group == "Treatment") %>%
ggplot(aes(x = amount)) +
geom_histogram(binwidth = 5, fill = "seagreen", alpha = 0.7) +
geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
labs(title = "Treatment Group: Donation Amounts",
    x = "Amount Donated", y = "Count") +
theme_minimal()
```

Both groups show right-skewed distributions of donation amounts. The dashed red lines represent the average amount given. There is no dramatic difference in the average donation size between treatment and control.


## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.

### Law of Large Numbers

_to do:  Make a plot like those on slide 43 from our first class and explain the plot to the reader. To do this, you will simulate 100,00 draws from the control distribution and 10,000 draws from the treatment distribution. You'll then calculate a vector of 10,000 differences, and then you'll plot the cumulative average of that vector of differences. Comment on whether the cumulative average approaches the true difference in means._

```{r}
# Set seed for reproducibility
set.seed(123)

# Simulate 10,000 draws from Bernoulli p=0.022 (treatment)
treatment_sim <- rbinom(10000, 1, 0.022)

# Simulate 10,000 draws from Bernoulli p=0.018 (control)
control_sim <- rbinom(10000, 1, 0.018)

# Compute difference at each point
diffs <- treatment_sim - control_sim

# Cumulative average of the differences
cum_avg <- cumsum(diffs) / seq_along(diffs)

# Plot
ggplot(data.frame(index = 1:10000, cum_avg = cum_avg), aes(x = index, y = cum_avg)) +
geom_line(color = "steelblue") +
geom_hline(yintercept = 0.004, linetype = "dashed", color = "red") +
labs(title = "Law of Large Numbers Simulation",
    x = "Number of Observations",
    y = "Cumulative Average of Differences") +
theme_minimal()
```

The plot shows the cumulative average of differences in donation probabilities converging toward 0.004 — the true difference (2.2% - 1.8%). This illustrates the **Law of Large Numbers**: as sample size grows, sample statistics converge to their true values.


### Central Limit Theorem

_to do: Make 4 histograms like those on slide 44 from our first class at sample sizes 50, 200, 500, and 1000 and explain these plots to the reader. To do this for a sample size of e.g. 50, take 50 draws from each of the control and treatment distributions, and calculate the average difference between those draws. Then repeat that process 999 more times so that you have 1000 averages. Plot the histogram of those averages. Comment on whether zero is in the "middle" of the distribution or whether it's in the "tail."_

```{r}
simulate_diffs <- function(n, reps = 1000, p_control = 0.018, p_treatment = 0.022) {
replicate(reps, {
    control <- rbinom(n, 1, p_control)
    treatment <- rbinom(n, 1, p_treatment)
    mean(treatment) - mean(control)
})
}

sizes <- c(50, 200, 500, 1000)
diffs_by_size <- lapply(sizes, simulate_diffs)

# Plot histograms
par(mfrow = c(2, 2))  # 2 rows x 2 columns layout

for (i in 1:4) {
hist(diffs_by_size[[i]], breaks = 30, main = paste("Sample size =", sizes[i]),
    xlab = "Mean difference (treatment - control)", col = "lightblue", border = "white")
abline(v = 0, col = "red", lwd = 2, lty = 2)
}
```

As sample size increases, the distribution of average differences becomes more concentrated around the true mean and more symmetric. With larger samples (like 1000), the distribution approximates a normal curve — demonstrating the **Central Limit Theorem**.




